Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-14 paper/intro.tex
second pass over intro/related

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 60s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below. 
\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    \stepcounter{enumi}
  \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    \begin{enumerate}[leftmargin=0pt]
    \item \label{wp-write}
      $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    \item \label{wp-let}
      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    \item \label{wp-read}
      $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \\
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.  We have split
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
\eqref{wp-assign} into three cases. In our notation,
$\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
most once,
$\aLoc$--$\cLoc$ range over shared memory references, and
$\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
include
$\aLoc$--$\cLoc$.\footnote{We wrote \ref{wp-read} using implication rather
  than substitution; the two formulations are equivalent under the assumption
  that each register is assigned at most once.}

@@

Existing approaches to sequential composition in the concurrent context
either assume exclusive access, as in concurrent separation logic
\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
altogether, as in the pomset model of
\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
dependencies and thus dramatically limits compiler optimization.  This leaves
open the question of how to apply logic to programs with data races.
To understand the solution, one must first understand the constraints imposed
by hardware and compilers.

@@

Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizations. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
  \begin{gather*}
    \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
    \PAR
    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=15,in=165]{wy1}{ry1}
        \rf[out=165,in=15]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the control dependency.

To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
\emph{Pomsets with Preconditions}, where events are labeled with logical
formulae.  Nontrivial preconditions are introduced by store actions (modeling
data dependencies) and conditionals (modeling control dependencies):
  \begin{gather*}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
      \end{tikzinline}}
  \end{gather*}
Preconditions are discharged by being ordered after a read:
  \begin{gather*}
    \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
$(0{=}s)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
which is why we do not assume the hypothesis $(0{=}r)$. Nonetheless, the precondition on
$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
\begin{gather*}
  % \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  % \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{\DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}

@@

Predicate transformers are a good fit to logical models of dependency calculation,
since both are concerned with preconditions, and how they are transformed by
sequential composition. Our first attempt is to associate a predicate transformer
with each pomset. We visualize this in diagrams by showing how $\bForm$ is transformed,
for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        \xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer from the write matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}.  For the reads, however,
\ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
$(x{=}r) \limplies \bForm$.  Instead, we use $(0{=}r) \limplies \bForm$,
reflecting the fact that $0$ may come from a concurrent write.  The
obligation to find a matching write is moved from the sequential semantics of
\emph{substitution} and \emph{implication} to the concurrent semantics of
\emph{fulfillment}.

@@

For the sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to the preconditions of
subsequent events, composing the predicate transformers.  (In subsequent
diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\[-1.5ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{left=of rx0}
      \po{rx0}{ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0ry0d}
      \xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
This model works for the sequentially consistent case, but needs to be
weakened for the relaxed case. The key observation of this paper is
that rather than working with one predicate transformer, we should
work with a \emph{family} of predicate transformers, indexed by sets
of events.

@@

For example, we can recover the expected semantics
for the above example by choosing
the predicate transformer which is independent of $(\DR x0)$
but dependent on $(\DR y0)$, which is the transformer
which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
  \begin{gather*}
    \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\[-2ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.2em]
        \xform{rx0d}{(0{=}r) \limplies \bForm}{}
        \event{rx0}{\DR{x}{0}}{right=of rx0d}
        \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
        \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
        \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0d}
        \xform{wz0i}{\bForm}{left=of rx0d}
        \po[out=20,in=170]{ry0}{wz0}
        \xo{rx0}{rx0d}
        \xo{ry0}{ry0d}
        \xo{rx0}{rx0ry0d}
        \xo{ry0}{rx0ry0d}
      \end{tikzinline}}
  \end{gather*}
As a sanity check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left, with intermediate step:
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \xform{rx0d}{(0{=}r) \limplies \bForm}{}
      \event{rx0}{\DR{x}{0}}{right=of rx0d}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
      \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
      \xform{wz0i}{\bForm}{left=of rx0d}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
or to the right, with intermediate step:
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{0}}{}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a monoid.

@@

In this paper, we show how pomsets with
preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined with
predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} to create a
compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, which satisfies many
  desiderata, but not all.
\item \S\ref{sec:arm} shows two approaches for efficient implementation on
  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
  uses an optimal lowering, but requires changes to the definitions of
  parallel and sequential composition.
\item \S\ref{sec:variants} generalizes the basic semantics of read and write
  to validate compiler optimizations.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
%The definitions in this paper have been formalized in Agda.
Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% efficient compilation to \armeight{}, which improves upon that of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% characterization of \armeight{}.

@@

\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.

>>>>

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 60s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below. 
\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    \stepcounter{enumi}
  \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    \begin{enumerate}[leftmargin=0pt]
    \item \label{wp-write}
      $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    \item \label{wp-let}
      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    \item \label{wp-read}
      $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \\
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.  We have split
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
\eqref{wp-assign} into three cases. In our notation,
$\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
most once,
$\aLoc$--$\cLoc$ range over shared memory references, and
$\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
include
$\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
  equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}

@@

Existing approaches to sequential composition in the concurrent context
either assume exclusive access, as in concurrent separation logic
\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
altogether, as in the pomset model of
\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
dependencies and thus dramatically limits compiler optimization.  This leaves
open the question of how to apply logic to racy programs without
overstraining the implementation.  To understand the solution, one must first
understand the constraints imposed by hardware and compilers.

@@

While syntactic dependency calculation suffices for hardware models, it is
not preserved by common compiler optimizations. For example, if we calculate
control dependencies syntactically, then there is a dependency from
$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
execution:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the dependency.

To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
\emph{Pomsets with Preconditions}, where events are labeled with logical
formulae.  Nontrivial preconditions are introduced by store actions (modeling
data dependencies) and conditionals (modeling control dependencies):
\begin{gather*}
  \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
    \end{tikzinline}}
\end{gather*}
Preconditions are discharged by being ordered after a read:
\begin{gather*}
  \tag{$\dagger$}\label{eq1}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}
Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
$(0{=}s)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
which is why we do not assume the hypothesis $(0{=}r)$. Nonetheless, the precondition on
$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
\begin{gather*}
  % \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  % \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{\DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}

@@

Predicate transformers are a good fit for logical models of dependency
calculation, since both are concerned with preconditions and how they are
transformed by sequential composition. Our first attempt is to associate a
predicate transformer with each pomset. We visualize this in diagrams by
showing how $\bForm$ is transformed, for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        \xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer from the write matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}.  For the reads, however,
\ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
$(x{=}r) \limplies \bForm$.  Instead, we use $(0{=}r) \limplies \bForm$,
reflecting the fact that $0$ may come from a concurrent write.  The
obligation to find a matching write is moved from the sequential semantics of
\emph{substitution} and \emph{implication} to the concurrent semantics of
\emph{fulfillment}.

@@

For a sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to the preconditions of
subsequent events, composing the predicate transformers.  (In subsequent
diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\[-1.5ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{left=of rx0}
      \po{rx0}{ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0ry0d}
      \xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
This model works for the sequentially consistent case, but needs to be
weakened for the relaxed case. The key observation of this paper is
that rather than working with one predicate transformer, we should
work with a \emph{family} of predicate transformers, indexed by sets
of events.

@@

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x0)$
but dependent on $(\DR y0)$, which is the transformer
which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\[-2ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.2em]
      \xform{rx0d}{(0{=}r) \limplies \bForm}{}
      \event{rx0}{\DR{x}{0}}{right=of rx0d}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
      \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0d}
      \xform{wz0i}{\bForm}{left=of rx0d}
      \po[out=20,in=170]{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
As a sanity check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left, with intermediate step:
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \xform{rx0d}{(0{=}r) \limplies \bForm}{}
      \event{rx0}{\DR{x}{0}}{right=of rx0d}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
      \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
      \xform{wz0i}{\bForm}{left=of rx0d}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
or to the right, with intermediate step:
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{0}}{}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a monoid.

@@

We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
be added to pomsets with
preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, which satisfies many
  desiderata, but not all.
\item \S\ref{sec:arm} shows two approaches for efficient implementation on
  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
  uses an optimal lowering, but requires changes to the definitions of
  parallel and sequential composition.
\item \S\ref{sec:variants} generalizes the basic semantics of read and write
  to validate compiler optimizations.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
% The definitions in this paper have been formalized in Agda.
Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% efficient compilation to \armeight{}, which improves upon that of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% characterization of \armeight{}.

@@

\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.  The use step-indexing to account for
loops; we expect that the same approach could be applied here.
